---
title: "DS202 lab 5"
author: "Andrew Fahmy"
date: "April 16, 2021"
output: html_document
---

```{r setup, include = FALSE}
    knitr::opts_chunk$set(echo = TRUE)
```

```{r}
    acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
    person <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```

# 1. Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?
```{r}
    library(tidyverse)
    names(acc)
    names(person)
    acc %>% group_by(DAY_WEEK) %>% summarise(n = n())
    acc %>% ggplot(aes(x = DAY_WEEK)) + geom_bar()
```

# 2. Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)
```{r}
    fatally_hurt <- person %>% filter(INJ_SEV == 4)
    #names(fatally_hurt)
    #head(fatally_hurt)
```


# 3. Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)
```{r}
    dangerous_vehicle <- fatally_hurt %>%
                             mutate(MAKE = ifelse(is.na(MAKE), 99, MAKE)) %>%
                             select(STATE, MAKE) %>%
                             group_by(STATE) %>%
                             count(MAKE) %>%
                             top_n(1)
    #head(dangerous_vehicle)
```
Accourding to the FARS manual, 99 is unknown make. We replace NA with 99 to get
rid of NA for the MAKE


# 4. Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)
```{r}
    #install.packages("maps")
    library(maps)
    library(readxl)
    states <- map_data("state")
    names(states)
    glcs_us <- read_xlsx("GLCs_US.xlsx")
    names(glcs_us)

    #data <- states %>% left_join(glcs_us, by = c("
    states %>% ggplot(aes(x = long, y = lat, group = group)) + geom_polygon()
```


# 5. Join the accident and person table (work out which variable(s) to use)
```{r}
    dat <- acc %>% inner_join(person, by = "ST_CASE")
    dim(dat)
    head(dat)
```


# 6. Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results and explain what you find.
```{r}
    # Day of week
    acc %>% group_by(DAY_WEEK) %>% summarise(n = n())
    acc %>% ggplot(aes(x = DAY_WEEK)) + geom_bar()
    # Hour of day
    acc %>% group_by(HOUR) %>% summarise(n = n())
    acc %>% filter(HOUR != 99) %>% ggplot(aes(x = HOUR)) + geom_bar()
    # Sex
    dat %>% group_by(SEX) %>% summarise(n = n())
    dat %>% filter(SEX <= 2) %>% ggplot(aes(x = SEX)) + geom_bar()
```


# 7. Now plot a choropleth map of the number of deaths on a county level. Also explain what you find.
```{r}

```


# 8. Is summer or winter more dangerous? Does this depend on states? Explore and explain.
```{r}
    acc %>% group_by(MONTH) %>% summarise(n = n())
    acc %>% ggplot(aes(x = MONTH)) + geom_bar()
```
